Future challenges for Research Software Engineering

### 2025-08-11 future_challenges_for_RSEng.tex
typo

So where do RSEs fit into this picture? Even today, we expect to see RSEs providing technical skills and input
to research activities in a range of different areas, including:
high-quality, robust architecture and design for research software;
seamless integration with research and data infrastructures;
development of software for novel and emerging architectures;
application of specialist numerical and statistical methods within software;
optimisation of codes for use on high-performance computing infrastructure;
and green computing - efficient implementations of algorithms/methods and use of hardware

@@

Besides the direct impact in research software and further research output,
we see a number of risks.
The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
We address this in \autoref{sec:complexity}.
Another risk is that the widespread use of AI tools will make community resources
such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible. Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more driven by institutions able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
As more and more research will be written by researchers assisted by AI,
RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
To address these risks, the RSE of the future could be not only closely overlooking every part of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
but also contributing to endangered community resources and infrastructure that will keep research accessible to the wider scientific community.

@@

\subsection{Impact of Complexity and Usability in Research Software}
\label{sec:complexity}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}

>>>>

So where do RSEs fit into this picture? Even today, we expect to see RSEs providing technical skills and input
to research activities in a range of different areas, including:
high-quality, robust architecture and design for research software;
seamless integration with research and data infrastructures;
development of software for novel and emerging architectures;
application of specialist numerical and statistical methods within software;
optimisation of codes for use on high-performance computing infrastructure;
and green computing - efficient implementations of algorithms/methods and use of hardware.

@@

Another risk of the widespread use of AI tools is that community resources
such as \eg\ Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.

@@

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}
